---
title: "`r paste0('Soil-fauna scenario report: ', params$scenario)`"
subtitle: "Millennial soil-carbon model coupled to soil-food-web fauna"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
output:
  pdf_document:
    toc: true
    number_sections: true
geometry: margin=1cm
params:
  scenario: "Earthworm"
  model: "millennial"
  project_root: ".."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5, dpi = 110)
library(tidyverse); library(knitr)

# Resolve project root so the report can be knit from anywhere.
root <- normalizePath(params$project_root, mustWork = TRUE)
knitr::opts_knit$set(root.dir = root)
```

```{r sources, include=FALSE}
setwd(root)
suppressPackageStartupMessages({
  library(deSolve); library(rootSolve); library(yaml); library(readxl)
})
source("R/climate_forcing.R"); source("R/spinup.R"); source("R/plot_ode_output.R")
source("R/setup.R");           source("R/compare_functions.R")
source("R/fit_animals.R");     source("R/dynamic_spinup.R")
source("R/report_metadata.R")

model    <- params$model
scenario <- params$scenario
```

```{r compute, include=FALSE}
# ------------------------------------------------------------
# Rebuild the scenario from the saved spun-up parameters so the report reflects
# EXACTLY the run that produced the saved states. If a saved spin-up exists we
# use its parms; otherwise we rebuild from the scenarios file.
# ------------------------------------------------------------
scen <- read_scenarios("Data/scenarios.xlsx")
spin_file <- sprintf("Data/spinup/%s_%s_baseline.rds", model, scenario)
have_spin <- file.exists(spin_file)
saved <- if (have_spin) load_spinup(spin_file) else NULL

# treatment + baseline setups (fresh); apply fitted animal params + saved parms
fitted <- if (file.exists("Results/fitted_animal_params.csv"))
  load_fitted_params("Results/fitted_animal_params.csv") else NULL

pair <- setup_scenario_pair(model, scen, scenario)
if (!is.null(fitted))
  pair$treatment <- apply_fitted_params(pair$treatment, fitted, model, scenario, verbose = FALSE)

# --- EQUILIBRIUM (constant-forcing) carbon, both arms ---
pair$baseline  <- spinup_equilibrium(pair$baseline,  verbose = FALSE)
pair$treatment <- spinup_equilibrium(pair$treatment,
                                     warm_start = pair$baseline$init_state_spin, verbose = FALSE)
eq_base <- pair$baseline$init_state_spin
eq_trt  <- pair$treatment$init_state_spin

# --- DIRECT effect: indirect (_pint + exudate) slopes zeroed ---
pair_d  <- zero_indirect_effects(pair, verbose = FALSE)
eq_dir  <- spinup_equilibrium(pair_d$treatment, warm_start = eq_base, verbose = FALSE)$init_state_spin

eff_total  <- compare_vectors(eq_trt, eq_base); eff_total$type  <- "total"
eff_direct <- compare_vectors(eq_dir, eq_base); eff_direct$type <- "direct"
eq_effect  <- rbind(eff_total, eff_direct)

# --- SPIN-UP (seasonal limit-cycle) baseline carbon ---
# The saved state is the t = 0 point of the limit cycle. To report the ANNUAL
# MEAN (the fairer number to compare with the equilibrium), integrate one
# seasonal year from the saved state and take the time-weighted (trapezoidal)
# mean of each pool over that year. `spin_t0` keeps the raw t = 0 snapshot for
# reference.
spin_t0   <- if (have_spin) saved$state else NULL
spin_mean <- NULL
if (have_spin) {
  y0    <- pair$baseline$init_state_spin
  y0[names(spin_t0)] <- spin_t0[names(y0)]          # start from the saved cycle state
  pb    <- pair$baseline$parms
  pb$climate_forcing <- make_climate_forcing(pb)     # full seasonal forcing
  yr    <- deSolve::ode(y = y0[names(pair$baseline$working_state)],
                        times = seq(0, 365, by = 1),
                        func = pair$baseline$wrapped_model, parms = pb)
  yr    <- as.data.frame(yr)
  pcols <- setdiff(names(yr), c("time", "mass_balance_check"))
  # time-weighted mean over the year (trapezoidal / uniform daily grid = colMeans)
  spin_mean <- setNames(
    vapply(pcols, function(cn) {
      v <- yr[[cn]]; tt <- yr$time
      sum(diff(tt) * (head(v, -1) + tail(v, -1)) / 2) / (max(tt) - min(tt))
    }, numeric(1)), pcols)
}

# --- OVER-SIMULATION effect from saved follow-up runs (add vs continue) ---
add_file  <- sprintf("Data/followup/%s_%s_add.rds", model, scenario)
ctrl_file <- sprintf("Data/followup/%s_%s_continue_baseline.rds", model, scenario)
have_follow <- file.exists(add_file) && file.exists(ctrl_file)
if (have_follow) {
  add  <- load_followup(model, scenario, "add")$out  |> as.data.frame()
  ctrl <- load_followup(model, scenario, "continue_baseline")$out |> as.data.frame()
  names(add)[1] <- names(ctrl)[1] <- "time"
  keep_p <- setdiff(intersect(names(add), names(ctrl)), c("time", "mass_balance_check"))
  tt <- intersect(add$time, ctrl$time)
  A <- add[match(tt, add$time),  keep_p, drop = FALSE]
  C <- ctrl[match(tt, ctrl$time), keep_p, drop = FALSE]
  sim_change <- data.frame(time = tt, A - C)                    # absolute change (g C m-2)
  sim_pct    <- data.frame(time = tt, 100 * (A - C) / C)        # percent change
  # end-of-simulation snapshot
  last <- nrow(sim_change)
  sim_end <- tibble(pool = keep_p,
                    abs_change = as.numeric(sim_change[last, keep_p]),
                    pct_change = as.numeric(sim_pct[last, keep_p]),
                    years = tt[last] / 365)
}

# helper: pretty pool label + soil-only filter
lab <- function(x) ifelse(x %in% names(pool_label), pool_label[x], x)
```

# Overview {#overview}

This report documents the **`r scenario`** scenario of the `r model` model so
co-authors can vet the results. All carbon stocks are in **`r UNIT_POOL`**;
all state-variable derivatives (dPool/dt) are in **`r UNIT_FLUX`**; one model
year = 365 days. Rate constants are **`r UNIT_RATE_DAY`** and annual NPP inputs
are **`r UNIT_NPP_ANNUAL`** unless noted otherwise in the parameter table.

```{r overview-status}
tibble(
  Item = c("Model", "Scenario", "Active pools",
           "Saved equilibrium", "Saved seasonal spin-up", "Saved follow-up (add/continue)"),
  Value = c(model, scenario,
            paste(names(pair$treatment$working_state), collapse = ", "),
            "recomputed in this report",
            ifelse(have_spin, basename(spin_file), "NOT FOUND"),
            ifelse(have_follow, "found", "NOT FOUND"))
) |> kable(caption = "Scenario setup and data availability.")
```

The **active animal(s)** in this scenario:

```{r active-animals}
act_an <- intersect(animal_pools, names(pair$treatment$working_state))
if (length(act_an))
  tibble(Animal = lab(act_an), Pool = act_an) |> kable() else
  cat("No animal pools active (pure baseline scenario).")
```

# Baseline carbon {#baseline}

## Equilibrium (constant-forcing) stocks

Equilibrium is the steady state under constant annual-mean forcing
(Temp = MAT, moisture = MAtheta). Units: **`r UNIT_POOL`**.

```{r baseline-eq-table}
tibble(Pool = lab(names(eq_base)), Raw = names(eq_base),
       `Equilibrium (g C m-2)` = as.numeric(eq_base)) |>
  filter(Raw %in% soil_pools) |>
  arrange(desc(`Equilibrium (g C m-2)`)) |>
  mutate(`Equilibrium (g C m-2)` = round(`Equilibrium (g C m-2)`, 2)) |>
  select(-Raw) |>
  kable(caption = "Baseline equilibrium carbon stocks (soil + plant-root pools).")

tot_eq <- sum(eq_base[intersect(soil_pools, names(eq_base))])
```

**Total baseline equilibrium soil + root carbon: `r round(tot_eq, 1)` `r UNIT_POOL`.**

## Seasonal spin-up stocks

The seasonal spin-up is the limit cycle under the full within-year forcing. The
values below are the **annual mean** of each pool over one year of that cycle
(time-weighted over the seasonal trajectory), which is the fair number to
compare with the equilibrium. Units: **`r UNIT_POOL`**.

```{r baseline-spin-table}
if (have_spin) {
  ss <- spin_mean[intersect(soil_pools, names(spin_mean))]
  tibble(Pool = lab(names(ss)), `Spin-up annual mean (g C m-2)` = round(as.numeric(ss), 2)) |>
    arrange(desc(`Spin-up annual mean (g C m-2)`)) |>
    kable(caption = "Baseline seasonal spin-up carbon (annual mean over the limit cycle, g C m-2).")
  tot_spin <- sum(spin_mean[intersect(soil_pools, names(spin_mean))])
} else {
  cat("No saved seasonal spin-up for this scenario (run Scripts/spinup_dynamic.R).")
}
```

`r if (have_spin) sprintf("**Total baseline seasonal spin-up soil + root carbon (annual mean): %.1f %s.**", tot_spin, UNIT_POOL) else ""`

```{r baseline-eq-vs-spin, fig.height=5, fig.cap="Baseline carbon by pool: equilibrium vs. seasonal spin-up annual mean (g C m-2, log scale)."}
if (have_spin) {
  common <- intersect(names(eq_base), names(spin_mean))
  common <- intersect(common, soil_pools)
  df <- bind_rows(
    tibble(Pool = lab(common), C = as.numeric(eq_base[common]),  kind = "Equilibrium"),
    tibble(Pool = lab(common), C = as.numeric(spin_mean[common]), kind = "Seasonal spin-up (annual mean)"))
  ggplot(df, aes(reorder(Pool, C), C, color = kind)) +
    geom_point(position = "dodge") + coord_flip() + scale_y_log10() +
    labs(x = NULL, y = "Carbon (g C m-2, log scale)", color = NULL) +
    theme_minimal(base_size = 11) + theme(legend.position = "top")
} else cat("(no spin-up to compare)")
```

# Animal effect at equilibrium {#eq-effect}

The animal effect is the treatment (animals present) minus baseline (no animals)
at equilibrium, reported as **absolute change (`r UNIT_POOL`)** and **percent
change**. *Total* = all pathways; *Direct* = feeding fluxes only (the indirect
`_pint` and exudation slopes set to zero).

```{r eq-effect-table}
eq_effect |>
  filter(!is.na(baseline), name %in% soil_pools) |>
  transmute(Pool = lab(name), Type = type,
            `Baseline (g C m-2)` = round(baseline, 2),
            `Treatment (g C m-2)` = round(treatment, 2),
            `Abs change (g C m-2)` = round(difference, 3),
            `% change` = round(percent_change, 3)) |>
  arrange(Pool, Type) |>
  kable(caption = "Equilibrium animal effect by pool (total and direct).")
```

```{r eq-effect-plot, fig.height=6, fig.cap="Equilibrium animal effect (% change vs. baseline) by pool; total vs. direct."}
eq_effect |>
  filter(!is.na(baseline), name %in% soil_pools) |>
  mutate(Pool = lab(name)) |>
  ggplot(aes(reorder(Pool, percent_change), percent_change, fill = type)) +
  geom_col(position = "dodge") + coord_flip() +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  labs(x = NULL, y = "Animal effect (% change vs. baseline)", fill = "Effect") +
  scale_fill_manual(values = c(total = "#1b7837", direct = "#762a83")) +
  theme_minimal(base_size = 11) + theme(legend.position = "top")
```

# Animal effect over the simulation {#sim-effect}

Change through the follow-up = animals-added run minus the time-matched
continued-baseline (no-animal) run, per pool. Absolute change in `r UNIT_POOL`;
percent relative to the continued baseline.

```{r sim-effect-table}
if (have_follow) {
  sim_end |>
    filter(pool %in% soil_pools) |>
    transmute(Pool = lab(pool),
              `Abs change (g C m-2)` = round(abs_change, 3),
              `% change` = round(pct_change, 3)) |>
    arrange(desc(abs(`Abs change (g C m-2)`))) |>
    kable(caption = sprintf("Animal effect at end of the %.0f-year follow-up.",
                            sim_end$years[1]))
} else cat("No saved follow-up runs (run Scripts/followup_analysis.R).")
```

```{r sim-effect-traj, fig.height=6, fig.cap="Animal effect trajectory (added - continued baseline) over the follow-up, by pool (g C m-2)."}
if (have_follow) {
  long <- sim_change |>
    pivot_longer(-time, names_to = "pool", values_to = "change") |>
    filter(pool %in% soil_pools) |>
    mutate(Pool = lab(pool), years = time / 365)
  ggplot(long, aes(years, change, colour = Pool)) +
    geom_hline(yintercept = 0, linewidth = 0.3, colour = "grey50") +
    geom_line(linewidth = 0.6) +
    labs(x = "Years after animals added", y = "Change in C (added - baseline, g C m-2)",
         colour = "Pool") +
    theme_minimal(base_size = 11)
} else cat("(no follow-up trajectory to plot)")
```

# Parameters unique to this scenario {#params}

Parameters whose value **differs from the model defaults** (common.yml +
`r paste0(model, '.yml')`). These are what make this scenario distinct;
shared defaults are omitted. Units and descriptions from the model metadata.

```{r unique-params}
defaults <- modifyList(yaml::read_yaml("config/common.yml"),
                       yaml::read_yaml(sprintf("config/%s.yml", model)))
p_used <- pair$treatment$parms
cand   <- intersect(names(p_used), names(defaults))
is_scalar_num <- function(x) is.numeric(x) && length(x) == 1 && is.finite(x)

diff_rows <- lapply(cand, function(nm) {
  a <- p_used[[nm]]; b <- defaults[[nm]]
  if (!is_scalar_num(a) || !is_scalar_num(b)) return(NULL)
  if (isTRUE(all.equal(a, b, tolerance = 1e-10))) return(NULL)
  info <- param_info(nm)
  tibble(Parameter = nm, Label = info$label, Unit = info$unit,
         Default = b, `Scenario value` = a, Description = info$desc)
})
uniq <- bind_rows(diff_rows)
if (nrow(uniq)) {
  uniq |> arrange(Parameter) |> kable(caption = "Parameters unique to this scenario (differ from defaults).")
} else {
  cat("No parameter overrides: this scenario uses the model defaults, differing only in which animal pools are active (see Overview).")
}
```

```{r animal-params}
# animal parameters for the active fauna (the biological definition of the treatment)
act_an <- intersect(animal_pools, names(pair$treatment$working_state))
if (length(act_an)) {
  # crude name match: parameters mentioning the animal or its known levers
  levers <- c(Earthworm = "earthworm", Detritivore = "detritiv",
              DetPredator = "detpredator", RootHerb = "rootherb|root_herb|exudate_slope")
  pat <- paste(levers[act_an], collapse = "|")
  anp <- names(p_used)[grepl(pat, names(p_used), ignore.case = TRUE)]
  anp <- anp[vapply(anp, function(n) is_scalar_num(p_used[[n]]), logical(1))]
  if (length(anp))
    tibble(Parameter = anp,
           Label = vapply(anp, function(n) param_info(n)$label, ""),
           Value = round(vapply(anp, function(n) p_used[[n]], 0), 8),
           Unit  = vapply(anp, function(n) param_info(n)$unit, ""),
           Description = vapply(anp, function(n) param_info(n)$desc, "")) |>
      kable(caption = "Animal parameters governing the active fauna in this scenario.")
}
```

# Forcing functions {#forcing}

The seasonal forcing supplies temperature, moisture, and three within-year
input-timing weights (each integrates to 1 over the year, so annual input =
allocation x annual NPP). The equilibrium uses the annual means
(Temp = MAT = `r p_used$MAT` deg C, moisture = MAtheta = `r p_used$MAtheta` m3 m-3).

```{r forcing-table}
forcing_info |>
  rename(Forcing = forcing, Unit = unit, Description = description) |>
  kable(caption = "Forcing functions supplied to the model.")
```

```{r forcing-plot, fig.height=6, fig.cap="Seasonal forcing over one year: temperature, moisture, and the three NPP input-timing weights."}
cf  <- make_climate_forcing(p_used)
doy <- seq(0, 365, by = 1)
fv  <- t(sapply(doy, cf))
fd  <- as.data.frame(fv); fd$doy <- doy
fd |>
  pivot_longer(-doy, names_to = "forcing", values_to = "value") |>
  left_join(forcing_info, by = "forcing") |>
  mutate(facet = paste0(forcing, " (", unit, ")")) |>
  ggplot(aes(doy, value)) + geom_line(linewidth = 0.6, colour = "#2166ac") +
  facet_wrap(~facet, scales = "free_y") +
  labs(x = "Day of year", y = NULL) + theme_minimal(base_size = 10)
```

# Reproducibility {#repro}

```{r repro}
tibble(Item = c("Report generated", "Model", "Scenario",
                "Saved spin-up file", "Saved follow-up files", "R version"),
       Value = c(format(Sys.time()), model, scenario,
                 ifelse(have_spin, spin_file, "none"),
                 ifelse(have_follow, paste(basename(add_file), basename(ctrl_file)), "none"),
                 R.version.string)) |>
  kable()
```

All stocks in `r UNIT_POOL`; derivatives in `r UNIT_FLUX`; rates in
`r UNIT_RATE_DAY`; annual NPP in `r UNIT_NPP_ANNUAL`. One model year = 365 days.
